---
title: "Using mascarade package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using mascarade package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height=5)
```

This is a vignette describing usage of `mascarade` to generate masks for clusters
on 2D dimensional reduction plots like UMAP or t-SNE.

### Package installation

The package can be installed from GitHub:

```{r eval=FALSE}
remotes::install_github("alserglab/mascarade")
```

### Loading necessary libraries

```{r}
library(mascarade)
library(data.table)
library(ggplot2)
```

### Example run

Loading example data from PBMC 3K processed with Seurat (see below for more details).

```{r}
data("exampleMascarade")
```

UMAP coordinates:

```{r}
head(exampleMascarade$dims)
```

Cluster annotations:
```{r}
head(exampleMascarade$clusters)
```

Expression table for several genes:
```{r}
head(exampleMascarade$features)
```

Let's plot this data:

```{r}
data <- data.table(exampleMascarade$dims, 
                   cluster=exampleMascarade$clusters,
                   exampleMascarade$features)
ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(aes(color=cluster)) + 
    coord_fixed() + 
    theme_classic()
                               
```

Now let's generate cluster masks:

```{r}
maskTable <- generateMask(dims=exampleMascarade$dims, 
                          clusters=exampleMascarade$clusters)
```

The `maskTable` is actually a table of cluster borders. 
A single cluster can have multiple connected parts, and
one a single part can contain multiple border lines (groups).

```{r}
head(maskTable)
```

Now we can use this table to draw the borders with `geom_path` (`group` column should be used as the group aesthetics):

```{r}
ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(aes(color=cluster)) + 
    geom_path(data=maskTable, aes(group=group)) +
    coord_fixed() + 
    theme_classic()
```

Or we can color the borders instead of points:

```{r}
ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(color="grey") + 
    geom_path(data=maskTable, aes(group=group, color=cluster), linewidth=1) +
    coord_fixed() + 
    theme_classic()
```

Now we can easily show association between cell types and 
expression of particular genes, such as GNLY being a good marker
for NK cells in this dataset.

```{r}
library(ggnewscale) # for having two color scales simultaneously
ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(aes(color=GNLY), size=0.5) +
    scale_color_gradient2(low = "#404040", high="red") + 
    new_scale_color() + 
    geom_path(data=maskTable, aes(group=group, color=cluster), linewidth=1) +
    coord_fixed() + 
    theme_classic()
```

We can focus on a single cluster too:

```{r}
ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(aes(color=GNLY), size=0.5) + 
    scale_color_gradient2(low = "#404040", high="red") + 
    geom_path(data=maskTable[cluster=="NK"], aes(group=group)) +
    coord_fixed() + 
    theme_classic()
```

However, for showing individual clusters another mode of `generateMask` can be preferred,
which processes clusters independently and allows overlaps between masks.
Such masks are more inclusive, which can be particularly important when your
clusters don't separate well.


```{r}
maskTable <- generateMask(dims=exampleMascarade$dims, 
                          clusters=exampleMascarade$clusters,
                          type = "independent")

ggplot(data, aes(x=UMAP_1, y=UMAP_2)) + 
    geom_point(aes(color=GNLY), size=0.5) + 
    scale_color_gradient2(low = "#404040", high="red") + 
    geom_path(data=maskTable[cluster=="NK"], aes(group=group)) +
    coord_fixed() + 
    theme_classic()
```

### Working with Seurat

For this part of the vignette you need `Seurat` and `SeuratData` packages (the latter can be installed with `remotes::install_github('satijalab/seurat-data')`).

Loading the example dataset:

```{r message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratData)
InstallData("pbmc3k")
LoadData("pbmc3k")

pbmc3k.final <- Seurat::UpdateSeuratObject(pbmc3k.final)
```

Let's plot some features:
```{r}
featureList <- c("MS4A1", "GNLY", "CD3E", "CD14")
FeaturePlot(pbmc3k.final, features=featureList)
```

Generate masks from UMAP data:

```{r}
maskTable <- generateMask(
    dims=Embeddings(pbmc3k.final, "umap"),
    clusters=pbmc3k.final$seurat_annotations)
```

Now we can plot the same features with borders (there will be some warnings due to the scale change):

```{r message=FALSE, warning=FALSE}
plots <- FeaturePlot(pbmc3k.final, features=featureList, combine = FALSE)

plots <- lapply(plots, `+`,
                list(
                    geom_path(data=maskTable, aes(x=UMAP_1, y=UMAP_2, group=group)),
                    # so that borders aren't cropped:
                    scale_x_continuous(expand = expansion(mult = 0.05)),
                    scale_y_continuous(expand = expansion(mult = 0.05))) 
                )

patchwork::wrap_plots(plots)
```
Works with t-SNE too:

```{r message=FALSE, warning=FALSE}
pbmc3k.final <- RunTSNE(pbmc3k.final)

maskTable <- generateMask(
    dims=Embeddings(pbmc3k.final, "tsne"),
    clusters=pbmc3k.final$seurat_annotations)

plots <- FeaturePlot(pbmc3k.final, features=featureList, combine = FALSE, reduction = "tsne")

plots <- lapply(plots, `+`,
                list(
                    geom_path(data=maskTable, aes(x=tSNE_1, y=tSNE_2, group=group)),
                    # so that borders aren't cropped:
                    scale_x_continuous(expand = expansion(mult = 0.05)),
                    scale_y_continuous(expand = expansion(mult = 0.05))) 
                )

patchwork::wrap_plots(plots)
```
In this case it can help to play with the parameters a bit: 
increase smoothing and decrease the density threshold.
WARNING: the sense and scaling of the parameters can change in the future versions.


```{r message=FALSE, warning=FALSE}
maskTable <- generateMask(
    dims=Embeddings(pbmc3k.final, "tsne"),
    clusters=pbmc3k.final$seurat_annotations,
    minDensity = 0.05, smoothSigma = 0.04)

plots <- FeaturePlot(pbmc3k.final, features=featureList, combine = FALSE, reduction = "tsne")

plots <- lapply(plots, `+`,
                list(
                    geom_path(data=maskTable, aes(x=tSNE_1, y=tSNE_2, group=group)),
                    # so that borders aren't cropped:
                    scale_x_continuous(expand = expansion(mult = 0.05)),
                    scale_y_continuous(expand = expansion(mult = 0.05))) 
                )

patchwork::wrap_plots(plots)
```
